Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics

### 2014-10-02 anderson-etal-blackswan-timeseries.tex
Edits to ms

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{lineno}
\usepackage{ragged2e}

@@

The Student-t Gompertz model is:
\begin{align*}
x_t &= \ln N_t\\
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align*}

\noindent where $x_t$ is the $\ln$ abundance at time $t$. The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependence if $b < 0$. The parameter $\lambda$ represents the expected population growth rate at $x_t = 0$. The process noise, $\epsilon_t$, is modelled as a Student-t distribution that is centered at $0$, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$. If $\nu$ is small ($\lesssim 10$) the distribution has much heavier tails than a normal distribution.  For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}). By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.42\textwidth]{nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes. Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations. Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order. Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
Taxonomic orders with three or fewer populations in panel a are omitted for space.}\label{fig:nu-coefs}
\end{center}
\end{figure}

@@

For the Gompertz, the mostly weakly-informative priors I'm using are:
\begin{align*}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\nu &\sim \mathrm{Exponential}(0.01)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (5).
\end{align*}

@@

We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma).
\end{align*}
\noindent
In addition to the parameters in the original Gompertz model, an additional parameter, $\phi$, is estimated that represents the relationship between of subsequent process noise residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = 1, \mathrm{max.} = 1)$.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
  density of the Bayesian priors for the base Gompertz model. From left to right: (1) per capita growth
  rate at $\ln$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
    Cauchy} (2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Exponential}(0.01)$. (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Normal}(0, 1)[-1, 1]$. Not shown is $b$, the density dependence
  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
  \label{fig:priors}
\end{center}
\end{figure}

>>>>

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{lineno}
\usepackage{ragged2e}
\usepackage{color}

@@

The heavy-tailed Gompertz model is:
\begin{equation*}
%\begin{align*}
\ln N_t = x_t = \mathrm{Student\mhyphen t}( {\color{red} \nu }, \lambda + b x_{t-1}, \sigma),
%\end{align*}
\end{equation*}

\noindent where $x_t$ is the $\ln$ abundance $N$ at time $t$. The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependence if $b < 0$. The parameter $\lambda$ represents the expected population growth rate at $x_t = 0$. The process noise is modelled as a Student-t distribution with a scale parameter of $\sigma_\mathrm{proc}$, and a degrees of freedom parameter of $\nu$ (highlighted in red). If $\nu$ is small ($\lesssim 10$) the distribution has much heavier tails than a normal distribution.  For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}). By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.42\textwidth]{nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes.
Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order.
Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
Taxonomic orders with three or fewer populations in panel a are omitted for space.
Red and orange colours highlight populations with a high probability of heavy-tailed dynamics.
}
\label{fig:nu-coefs}
\end{center}
\end{figure}

@@

For the Gompertz, the mostly weakly-informative priors I'm using are:
\begin{align*}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\nu &\sim \mathrm{Exponential}(0.01)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (2.5).
\end{align*}

@@

We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma).
\end{align*}
\noindent
In addition to the parameters in the original Gompertz model, an additional parameter, $\phi$, is estimated that represents the relationship between of subsequent process noise residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
  density of the Bayesian priors for the base Gompertz model. From left to right: (1) per capita growth
  rate at $\ln$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
    Cauchy} (2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Exponential}(0.01)$. (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence
  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
  \label{fig:priors}
\end{center}
\end{figure}
